\vspace{-1cm}\normalsize{\textbf{Fairly Random: Impact of Winning the Toss on the Probability of Winning}}

### 2015-09-19 write_up/cricket.tex
fixing intro

setwd(paste0(githubdir, "/cricket-stats/write_up/"))	
tools::texi2dvi("cricket.tex", pdf=TRUE, clean=TRUE)	
setwd(paste0(githubdir, "/cricket-stats"))

@@

In nearly all cricket matches, it is claimed that there is a clear advantage to bowling (batting) first. The advantage is pointed to by commentators, by team captains in the pre-toss interview, and by the captain of the losing team in the post-match interview. This isn't to say that there is actually an advantage to winning the toss, or that in lies in what is claimed, or that teams are able to successfully exploit it. For it may be impossible to predict well in advance the advantage of bowling or batting first. Or it may be that teams squander the potential advantage by using bad heuristics to choose what they do.

The opportunity to choose whether to bowl or bat first is decided by a coin toss. While this method of granting advatange is fair on average, the system isn't fair in any one game. At first glance, the imbalance seems inevitable. After all, someone has to bat first. One can, however, devise a baseball like system where short innings are interspersed. If that violates the nature of the game too much, one can easily create pitches that don't deteriorate appreciably over the course of a game. Or, one can come up with an estimate of the advantage and adjust scores accordingly (something akin to an adjustment issued when matches are shortened due to rain).

@@

In limited overs cricket, a minimum number of overs must be bowled to establish a result. In a one-day match, for instance, each side must bat at least 20 overs for a result to be declared. In 706 matches, or roughly 1.7\% of the remaining matches, not enough overs were bowled to get a result. We again exclude these matches from our analysis. This leaves us with 39,672 matches. We analyse these data.

@@

The conventional wisdom among the lay cricket followers is that toss grants the greatest advantage in multi-day affairs like tests and first class matches, followed by day long affairs, and Twenty20s respectively. And there is good dose of common sense behind the conventional wisdom. Pitches invariably deteriorate over multiple days and batting last in a test match is often challenging. The pitch deteriorates far less over the course of the day, or in case of Twenty20s -- a few hours. And indeed unlimited over matches provide the most consistent advantage -- the average advantage over FC and test matches is north of 2.2\% (see Figure~\ref{fig:type}). But the heftiest advtange is in domestic one-day matches (List A), approximately 4\%. The number for ODIs, however, is a bit puzzling -- there is a slight disadvantage to winning the toss. That can best be ascribed to choosing badly.

>>>>

setwd(paste0(githubdir, "/cricket-stats/write_up/"))	
tools::texi2dvi("cricket.tex", pdf=TRUE, clean=TRUE)	
setwd(paste0(githubdir, "/cricket-stats/"))

@@

In nearly all cricket matches, it is claimed that there is a clear advantage to bowling (batting) first. The advantage is pointed to by commentators, by the team captains in the pre-toss interview, and by the captain of the losing team in the post-match interview. And to wrest this said advantage, one merely needs to pick the side of the coin that will be left facing the sky after it is tossed. And while this method of granting advantage is fair on average, the system isn't fair in any one game.

At first glance, the imbalance seems inevitable. After all, someone has to bat first. One can, however, devise a baseball like system where short innings are interspersed. If that violates the nature of the game too much, one can easily create pitches that don't deteriorate appreciably over the course of a game. Or, one can come up with an estimate of the advantage and adjust scores accordingly (something akin to an adjustment issued when matches are shortened due to rain).

None of this to say that there is actually an advantage to winning the toss, or, that teams are able to successfully exploit the potential advantage. For it may be impossible to predict well in advance the advantage of bowling or batting first. Or, it may be that teams squander the potential advantage by using bad heuristics to choose what they do.

@@

In limited overs cricket, a minimum number of overs must be bowled to establish a result. In a one-day match, for instance, each side must bat at least 20 overs for a result to be declared. In 706 matches, or roughly 1.7\% of the remaining matches, not enough overs were bowled to get a result. We again exclude these matches from our analysis. This leaves us with 39,672 matches. We analyze these data.

@@

The conventional wisdom among the lay cricket followers is that toss grants the greatest advantage in multi-day affairs like tests and first class matches, followed by day long affairs, and Twenty20s respectively. And there is good dose of common sense behind the conventional wisdom. Pitches invariably deteriorate over multiple days and batting last in a test match is often challenging. The pitch deteriorates far less over the course of the day, or in case of Twenty20s -- a few hours. And indeed unlimited over matches provide the most consistent advantage -- the average advantage over FC and test matches is north of 2.2\% (see Figure~\ref{fig:type}). But the heftiest advantage is in domestic one-day matches (List A), approximately 4\%. The number for ODIs, however, is a bit puzzling -- there is a slight disadvantage to winning the toss. That can best be ascribed to choosing badly.
